---
title: "DSC1105 | FA4"
author: "Baybayon, Darlyn Antoinette"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(cowplot)
})
```

1. Using the Mortality by Latitude data, make a plot of mortality index against mean average temperature. Is it hollow up or hollow down? Try to identify a transformation of one of the variables that will straighten out the relationship, and make a plot of the residuals to check for any remaining patterns.

```{r}
data <- read.csv("mortality_by_latitude.csv")
data

ggplot(data, aes(x= temperature, y=mortality_index)) + geom_point() +
  stat_smooth(method = "lm", formula = y ~ x,se=FALSE) +
  labs(title= "Mortality Index vs. Temperature")

ggplot(data, aes(x = temperature, y = log(mortality_index))) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(title="Log(Mortality Index) vs. Temperature")


model_log <- lm(log(mortality_index) ~ temperature, data = data)


ggplot(data, aes(x = temperature, y = resid(model_log))) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(y = "Residuals", x = "Temperature", title = "Residual Plot")

```




2. Using the same subset of the diamonds dataset, make a plot of log price as a function of carat with a loess smoother. Try several values for the span and degree arguments and comment briefly about your choice.



```{r}
diamond_plot <- ggplot(subset(diamonds, clarity == "VS1"), aes(x=carat, y=log(price))) +
  geom_point(size=0.5) 
  
diamond_plot + geom_smooth(method = "loess", se=FALSE, formula = y~x, span= 0.2, color="red") + 
labs(title="Span=0.2")
diamond_plot + geom_smooth(method = "loess", se=FALSE, formula = y~x, span= 0.5, color="red") + 
labs(title="Span=0.5")
diamond_plot + geom_smooth(method = "loess", se=FALSE, formula = y~x, span= 0.8, color="red") +
labs(title="Span=0.8")

```

I tested three spans, 0.2, 0.5, and 0.8. 0.2 follows a jagged line, 0.5 is comparably smoother while still showing the trends on the plot, and 0.8 is the smoothest. Among these values, 0.5 may be the best fit.



3. Compare the fit of the loess smoother to the fit of the polynomial + step function regression using a plot of the residuals in the two models. Which one is more faithful to the data? 


```{r}

# POLY STEP
step = function(x, step_position) {
    return(ifelse(x >= step_position, 1, 0))
}
lm_poly_step <- lm(log(price) ~ poly(carat, 3) + step(carat, c(0.3, 0.5, 1, 1.5, 2)), 
                   data = diamonds, subset = clarity == "VS1")

diamonds_poly_step <- subset(diamonds, clarity == "VS1") %>%
  mutate(poly_step_fitted = predict(lm_poly_step),
         poly_step_residual = log(price) - poly_step_fitted)


# LOESS
loess_fit <- loess(log(price) ~ carat, data = subset(diamonds, clarity == "VS1"), span = 0.2)

diamonds_loess <- subset(diamonds, clarity == "VS1") %>%
  mutate(loess_fitted = predict(loess_fit),
         loess_residual = log(price) - loess_fitted)

step_poly_plot <- ggplot(diamonds_poly_step, aes(x = carat, y = poly_step_residual)) +
  geom_point(size=0.75) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggtitle("Residuals: Poly + Step Function Regression")

loess_plot <- ggplot(diamonds_loess, aes(x = carat, y = loess_residual)) +
  geom_point(size=0.75) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggtitle("Residuals: Loess Smoother")


plot_grid(step_poly_plot, loess_plot)


```


Both models show residuals scattered around zero without strong patterns. Between these two, loess is more likely to follow the trends of the data smoothly compared to the poly + step function regression.
