---
title: "FA1"
author: "Baybayon, Darlyn Antoinette"
mainfont: SourceSansPro
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
knitr::opts_chunk$set(fig.width=8, out.width="77%", fig.height=6, out.height="45%")
```


```{r}
library(ggplot2)
library(magrittr)
library(stringr)
library(gridExtra)
dataset <- read.csv("cytof_one_experiment.csv")

save(dataset, file="my_work_space.RData")
```

##### ECDF plot for GranzymeB

```{r}
granzymeB <- dataset$GranzymeB
ggplot(dataset, aes(x = granzymeB)) + stat_ecdf() + xlab("GranzymeB") + ylab("percent")
```


This plot shows a slow increase at lower values, sharp rise in the middle. This means that most of the values in this column are concentrated in the middle and less extreme lower values.

##### Histogram plot for GranzymeB

```{r}
ggplot(dataset, aes(x = granzymeB)) +geom_histogram(binwidth = 1) + 
  xlab("GranzymeB") + ylab("percent")

```

The histogram for granzymeB is skewed to the left which means that most of the values in this column are large and there are less lower values.

##### Q-Q plot for CD2 and CD4

```{r}
cd2 <- dataset$CD2
cd4 <- dataset$CD4


qqplot(cd2, cd4,  main = "CD2 vs. CD4") 
```

This plot deviates significantly from a straight diagonal line which implies that the quantiles do not have a similar distribution.Towards the right, CD4 values increase sharply compared to CD2 which suggests that it is more skewed to the right. 



